---
title: "Gene Activation Analysis: Benchmark of NB-act using oncogenes"
output: github_document
---

# Introduction:
This analysis explores the performance of nb-act method for detecting activation outliers, that is, 
genes that are usually inactive in most of the samples (very low expression) and become active in a few samples (high expression). 
For benchmarking purposes, the method will be applied to the activation of oncogenes in patients with cancer.

In the normal state, oncogenes are typically involved in promoting cell growth, division, and survival. When mutated or overexpressed, can potentially cause normal cells become cancerous.
Therefore, we can expect to observe a high number of activated oncogenes in cancer patients. 

Specifically, we will use the Munich Leukemia Laboratory containing 3760 blood samples of male and female individuals with leukemia. 


## Load libraries, data and functions:  
```{r, echo = TRUE, message = FALSE}
library(dplyr)
library(SummarizedExperiment)
library(tibble)
library(tidyr)
library(ggplot2)
library(patchwork)
library(BiocGenerics)
```


```{r, echo = TRUE, results = FALSE, message = FALSE}
source("/data/nasif12/home_if12/l_araiz/workspace/gene_activation/scripts/function/plots_functions.R")
source("/data/nasif12/home_if12/l_araiz/workspace/gene_activation/scripts/function/help_functions.R")
source("/data/nasif12/home_if12/l_araiz/workspace/gene_activation/scripts/function/nb_act.R")
```


## Load data
```{r}
MLL <- readRDS("/s/project/mll/drop_2023feb/processed_results/aberrant_expression/v33b/outrider/leukemia_14group/ods_unfitted.Rds")
MLL <- OUTRIDER::estimateSizeFactors(MLL)


oncogenes_df <- read.table("/data/nasif12/home_if12/l_araiz/workspace/gene_act/oncogenes.tsv", sep = "\t", row.names = 1)
```


## Pre-processing

Select those genes that are rarely expressed (< 1 RPKM in > 95% of the samples). We would expect oncogenes to be rarely expressed, as there are usually inactive but become active in some cancer patients. 

```{r}
#Filter for rarely expressed genes (6401 genes)
rare_MLL <- filterRarely(MLL)

# There are 318 oncogenes
nrow(oncogenes_df)

# 314 of them are present in the MLL dataset
length(intersect(sub("\\..*", "", rownames(oncogenes_df)), sub("\\..*", "", rownames(rare_MLL$rpkm))))

# 62 of them are rarely expressed. 
length(intersect(sub("\\..*", "", rownames(oncogenes_df)), sub("\\..*", "", rownames(rare_MLL$rarely_exp_rpkm))))
```

For benchmarking purposes, let's see if we can find a subset of samples with more rarely expressed oncogenes. 

We can have a subset with 67 instead of 62 (not much better).
```{r}
# Load your RPKM matrix and oncogene list
# rpkm: rows are genes, columns are samples
# oncogenes: vector of oncogene IDs (e.g., ENSG000xxxx)

# Define the number of samples you want to select
subset_size <- 500 # Adjust as needed

# Run the optimization
result <- findOptimalSubset(rare_MLL$rpkm, oncogenes_df, subset_size)

# Extract the best subset of samples
best_samples <- result$best_subset
max_rare_genes <- result$max_rare_genes

cat("Best subset includes", max_rare_genes, "rarely expressed oncogenes.\n")
```


```{r}
# Just to check
counts_MLL <- assays(MLL)$counts
subset_counts_MLL <- counts_MLL[, best_samples]

rare_MLL_subset <- filterRarely(MLL, subset_counts = subset_counts_MLL)

length(intersect(sub("\\..*", "", rownames(oncogenes_df)), sub("\\..*", "", rownames(rare_MLL_subset$rarely_exp_rpkm))))
```


## Parameter exploration
```{r,  echo = TRUE, results = FALSE, message = FALSE}
# Parameter values to test 
threshold_vals <- c(0.1, 1, 5, 10, 50, 100)
theta_vals <- c(0.1, 1, 10, 100)

results_long_all_MLL <- exploreParamsOnco(MLL, rare_MLL_subset$rarely_exp_rpkm, theta_vals = theta_vals, threshold_vals = threshold_vals, interest_list = rownames(oncogenes_df))
```

```{r curve MLL onco}
results_long_all_MLL$results_long_all_mini <- results_long_all_MLL$results_long_all[which(results_long_all_MLL$results_long_all$interest == TRUE),]
curvePlot(results_long_all_MLL$results_long_all_mini, title = "Cumulative sum of oncogenes")
```

```{r N50 MLL_onco}
n50Plot(results_long_all_MLL, plot_type = "line")
```

In this case, low values for threshold and/or theta work pretty bad (= 0.1, 1). Best performance is achieved with very high values for both threshold (making it more conservative) and theta (making it less conservative). This behaviour differs from the pattern that we were observing in the Y-chr benchmark. 

Reasons might be that activated oncogenes have a higher expression (in absolute terms) than Y-chr genes, especially if the Y-chr gene is not highly expressed in the tissue. That is, a better signal to noise ratio. 
As a consequence, a higher threshold for describing "inactive" genes might remove part of this "noise". 


# CONCLUSION:

Y-chr and oncogenes benchmark present different patterns in terms of optimal parameters for modeling the distribution of inactive genes. 
In the case of Y-chr (both in GTEx and MLL), lower values of threshold (0.1, 1) and theta (<= 10) achieve a better performance. In the case of Oncogenes, higher values of threshold and theta are better. 

In general, oncogenes do not necessarily appear in the first positions when ranking outliers according to pvalue. Moreover, defining all oncogenes in all samples as positives might be overstimating the number of positive genes, as not every oncogene is expected to be activated in every patient. For this reasons, the oncogene approach is less suited for benchmarking purposes. 

1) Optimal parameters may depend on the type of gene that is being taken as a reference, rather than the dataset (Y-chr in GTEx and MLL is more similar than y-chr MLL and oncogenes MLL)

2) Y-chr works better than oncogenes. 


